Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-06-13 paper/document.tex
Fix LaTeX warnings.

\IEEEpeerreviewmaketitle

@@

Consider GitHub graphs~\cite{github-graphs}, visualizations that are easily available to open source software developers. GitHub's graphs use line and commit information exclusively. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the graph to be able to show details of other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort \cite{Hindle:2008:LCT:1370750.1370773}.

@@

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject of study due to its popularity. Grechanik \textit{et al.} \cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.} \cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.} \cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found what APIs are popular and if they were used in a framework-like manner. Dyer \textit{et al.} \cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and the adoption rate of new features over time. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.} \cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.} \cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies \cite{6980224} or files \cite{6650525, 7332421,voinea2006} over time.

Tools also exist which track changes over time in a more complex manner. Kim \textit{et al.} \cite{Kim:2009:DRS:1555001.1555046} built a tool called Logical Structural Diff (LSdiff) which examines code changes and produces logical rules to represent them. Changes which do not match the rules are then pointed out to the developer. Reiss and Tarvo created an IDE which allows users to see the history of files and changes to lines of code from each author as they are editing a project \cite{6650521}. Van Hees and Hage use a voronoi treemap to visualize changes in class and method structure \cite{7332410}. Servant and Jones created CHRONOS, a tool which creates a timeline of changes to a selection of code as well as the ability to query those changes \cite{6650547}.

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.} \cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files \cite{6980223}. This allowed testing practices to be more closely examined.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.} \cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. They found that developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.} \cite{lee2013} developed a tool for visualizing the branching structure of Git repositories. This was used to analyze concurrent workflows in popular open source repositories. Elsen also created a visualization for branches in Git, with the ability to expand branches to view the directory structure of a project at that point \cite{6650522}.   Shrestha \textit{et al.} made a visualization for the geographic location of a commit and \cite{6650532} and Minelli \textit{et al.} \cite{6980226} looked at the real time changes developers made to a repository.

@@

Some visualization tools are used for managing or making decisions about software rather than study the changes taking place. The service EasyBi \cite{EasyBi} provides additional visualizations for Git repositories for the purpose of business decisions. EasyBi works by taking the git log file and generating charts and graphs from it. It tries to solve the issue of summarizing many commits and has greater controls for granularity and authors but still operates on commits and line differences. The Gitinspector project \cite{Gitinspector} is an open source tool for the statistical analysis of Git repositories. It provides an advanced timeline view of changes per author and cumulative statistics but again relies upon the line differences to summarize the work of each author.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section \ref{sec:viz}.

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. When a programmer makes a non-trivial change in code in a file, that difference will be reflected in the AST. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no affect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by using calling \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure \ref{fig:commits}. Since we compared each commit to its parent, as reported by Git, we were able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure \ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

@@

Knowing type and author information enables other kinds of visualizations. For example, Figure \ref{fig:chord} is a chord diagram which shows connections between authors and types. The wider the arc around the circumference of the circle, the more often that entity is connected to other entities. For example, ``parrt'' is the top contributor, with changes covering a large number of types. Determining author and type information is made possible by AST differencing.

@@

When managing group projects in an introductory software engineering course, it is very important to be able to see the contributions that students are making to the group repository to judge learning, progress, and collaboration. When one of the authors was supervising 4-5 groups of students, it became infeasible to look through all of the commits of each student. Students may take the wrong approach in a certain piece of code that can become ingrained in the project and very difficult to change. Some groups had problems with group members who did not complete their assigned work or made meaningless contributions. Issues like this can be difficult to detect and prevent. TypeV was used by the author to quickly summarize the work done by each group member, organized by the types they had been working on. This can show when intervention or guidance is required, without the large time commitment of pouring through commits.

Bug triaging for large open source projects is a very time consuming task \cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.

>>>>

\IEEEpeerreviewmaketitle%

@@

Consider GitHub graphs~\cite{github-graphs}, visualizations that are easily available to open source software developers. GitHub's graphs use line and commit information exclusively. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the graph to be able to show details of other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort~\cite{Hindle:2008:LCT:1370750.1370773}.

@@

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in Section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject of study due to its popularity. Grechanik \textit{et al.}~\cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.}~\cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.}~\cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found what APIs are popular and if they were used in a framework-like manner. Dyer \textit{et al.}~\cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and the adoption rate of new features over time. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421,voinea2006} over time.

Tools also exist which track changes over time in a more complex manner. Kim \textit{et al.}~\cite{Kim:2009:DRS:1555001.1555046} built a tool called Logical Structural Diff (LSdiff) which examines code changes and produces logical rules to represent them. Changes which do not match the rules are then pointed out to the developer. Reiss and Tarvo created an IDE which allows users to see the history of files and changes to lines of code from each author as they are editing a project~\cite{6650521}. Van Hees and Hage use a voronoi treemap to visualize changes in class and method structure~\cite{7332410}. Servant and Jones created CHRONOS, a tool which creates a timeline of changes to a selection of code as well as the ability to query those changes~\cite{6650547}.

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.}~\cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. They found that developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.}~\cite{lee2013} developed a tool for visualizing the branching structure of Git repositories. This was used to analyze concurrent workflows in popular open source repositories. Elsen also created a visualization for branches in Git, with the ability to expand branches to view the directory structure of a project at that point~\cite{6650522}.   Shrestha \textit{et al.} made a visualization for the geographic location of a commit and~\cite{6650532} and Minelli \textit{et al.}~\cite{6980226} looked at the real time changes developers made to a repository.

@@

Some visualization tools are used for managing or making decisions about software rather than study the changes taking place. The service EasyBi~\cite{EasyBi} provides additional visualizations for Git repositories for the purpose of business decisions. EasyBi works by taking the git log file and generating charts and graphs from it. It tries to solve the issue of summarizing many commits and has greater controls for granularity and authors but still operates on commits and line differences. The Gitinspector project~\cite{Gitinspector} is an open source tool for the statistical analysis of Git repositories. It provides an advanced timeline view of changes per author and cumulative statistics but again relies upon the line differences to summarize the work of each author.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section~\ref{sec:viz}.

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. When a programmer makes a non-trivial change in code in a file, that difference will be reflected in the AST\@. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no affect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by using calling \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure~\ref{fig:commits}. Since we compared each commit to its parent, as reported by Git, we were able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure~\ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

@@

Knowing type and author information enables other kinds of visualizations. For example, Figure~\ref{fig:chord} is a chord diagram which shows connections between authors and types. The wider the arc around the circumference of the circle, the more often that entity is connected to other entities. For example, ``parrt'' is the top contributor, with changes covering a large number of types. Determining author and type information is made possible by AST differencing.

@@

When managing group projects in an introductory software engineering course, it is very important to be able to see the contributions that students are making to the group repository to judge learning, progress, and collaboration. When one of the authors was supervising 4--5 groups of students, it became infeasible to look through all of the commits of each student. Students may take the wrong approach in a certain piece of code that can become ingrained in the project and very difficult to change. Some groups had problems with group members who did not complete their assigned work or made meaningless contributions. Issues like this can be difficult to detect and prevent. TypeV was used by the author to quickly summarize the work done by each group member, organized by the types they had been working on. This can show when intervention or guidance is required, without the large time commitment of pouring through commits.

Bug triaging for large open source projects is a very time consuming task~\cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.
